---
title: "sentiment analysis NLP"
author: "LANGAT ERICK"
date: "2023-08-22"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **What is sentiment analysis?**

Sentiment analysis is the computational task of automatically determining what feelings a writer is expressing in text. Sentiment is often framed as a binary distinction (positive vs. negative), but it can also be a more fine-grained, like identifying the specific emotion an author is expressing (like fear, joy or anger).

Sentiment analysis is used for many applications, especially in business intelligence. Some examples of applications for sentiment analysis include:

-   Analyzing the social media discussion around a certain topic

-   Evaluating survey responses

-   Determining whether product reviews are positive or negative

Sentiment analysis is not perfect, and as with any automatic analysis of language, you will have errors in your results. It also cannot tell you why a writer is feeling a certain way. However, it can be useful to quickly summarize some qualities of text, especially if you have so much text that a human reader cannot analyze all of it.

```{r}
# This R program is designed and test for whatsapp chats(group or individual). 
#You can export the chats from whatsapp to your email ID and download it to your PC and work on it.
# It is defined by the kaggle/python docker image: https://github.com/kaggle/docker-python
library(bookdown)
library(tinytex)
library(ggplot2)
library(lubridate)
library(scales)
library(reshape2)
library(tm)
library(SnowballC)
library(wordcloud)
library(RColorBrewer)
library(stringr)
library(syuzhet) 
library(dplyr ) 
suppressPackageStartupMessages(require(tidyverse))
```

```{r}
#get the data from whatsapp chat 
text <- readLines("C:/Users/langa/OneDrive/Desktop/R PROGRAMMING PRACTICE/GBcomments.csv", n = 10000)
dim(text)
head(text)
```

```{r}
#Convert the file to UTF-8 encoding
s2 <- iconv(text, "UTF-8", "ASCII", sub = "")
text <- s2
head(s2)
```

```{r}
#let us create the corpus
docs <- Corpus(VectorSource(text))
head(docs)
```

```{r}
#clean our chat data
trans <- content_transformer(function (x , pattern ) gsub(pattern, " ", x))
docs <- tm_map(docs, trans, "/")
docs <- tm_map(docs, trans, "@")
docs <- tm_map(docs, trans, "\\|")
docs <- tm_map(docs, content_transformer(tolower))
docs <- tm_map(docs, removeNumbers)
docs <- tm_map(docs, removeWords, stopwords("english"))
docs <- tm_map(docs, removePunctuation)
docs <- tm_map(docs, stripWhitespace)
docs <- tm_map(docs, stemDocument)

```

```{r}
#create the document term matrix
dtm <- TermDocumentMatrix(docs)
mat <- as.matrix(dtm)
v <- sort(rowSums(mat),decreasing=TRUE)
```

```{r}
#Data frame
data <- data.frame(word = names(v),freq=v)
head(data)
```

```{r}
#generate the wordcloud
set.seed(1056)

wordcloud(words = data$word, freq = data$freq, min.freq = 1,
          max.words=200, random.order=FALSE, rot.per=0.35,
          colors=brewer.pal(8, "Dark2"))

```

```{r}
#fetch sentiment words from texts
Sentiment <- get_nrc_sentiment(as.character(text))
head(Sentiment)
text <- cbind(text,Sentiment)
head(text)

```

```{r}
#count the sentiment words by category
TotalSentiment <- data.frame(colSums(text[,c(2:11)]))
names(TotalSentiment) <- "count"
TotalSentiment <- cbind("sentiment" = rownames(TotalSentiment), TotalSentiment)
rownames(TotalSentiment) <- NULL
```

```{r}
#total sentiment score of all texts
ggplot(data = TotalSentiment, aes(x = sentiment, y = count)) +
  geom_bar(aes(fill = sentiment), stat = "identity") +
  theme(legend.position = "none") +
  xlab("Sentiment") + ylab("Total Count") + ggtitle("Total Sentiment Score")


```
